---
title: "Strawberries1"
author: "Keyu Feng"
date: "2024-10-05"
output: pdf_document
---

#Set up
```{r}
#Install the library
library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
setwd("C:/Users/16597/Downloads")
options(echo = FALSE, digits = 3, 
        scipen = 999, warn = FALSE, message = FALSE)

#Quick overview of the data
strawberry <- read_csv("strawberries25_v3.csv", col_names = TRUE)
glimpse(strawberry)

```

#Remove columns with a single value in all rows
```{r}
drop_one_value_col <- function(df){   ## takes whole dataframe
drop <- NULL  

## test each column for a single value
for(i in 1:dim(df)[2]){     
if((df |> distinct(df[,i]) |> count()) == 1){
drop = c(drop, i)
} }

## report the result -- names of columns dropped
## consider using the column content for labels 
## or headers 

if(is.null(drop)){return("none")}else{

   print("Columns dropped:")
   print(colnames(df)[drop])
   strawberry <- df[, -1*drop]
   }
}


## use the function

strawberry <- drop_one_value_col(strawberry)

drop_one_value_col(strawberry)
```

#Load all states data instead of just California
```{r}
# Remove the California-specific filter to explore all states
all_states <- strawberry

# Look at the unique values in the "Program" column for all states
unique(all_states$Program)

# Split the data into CENSUS and SURVEY groups
all_states_census <- all_states |> filter(Program == "CENSUS")
all_states_survey <- all_states |> filter(Program == "SURVEY")

# Select specific columns (Year, Period, Data Item, Value) for SURVEY data
all_states_survey <- all_states |> select(Year, Period, `Data Item`, Value)

```

#Split columns from sorted data
Looking at the data, I find that after splitting the columns, some information are false leading, my next goal is filling all valuables to their best-fit columns.
```{r}
#Replace ' - ' (hyphen with spaces) with a comma.
strawberry <- strawberry |>
  mutate(`Data Item` = str_replace_all(`Data Item`, " - ", ","))

#Split 'Data Item' into 4 columns
strawberry <- strawberry |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = ",",
                         names = c("Fruit",
                                 "Category",
                                 "Item",
                                 "Metric"),
                         too_many = "merge",
                         too_few = "align_start"
                       )
#Remove 'measured in' to metric columns
strawberry <- strawberry |>
  mutate(Metric = ifelse(grepl("MEASURED IN", Item), Item, Metric), # Move the 'Item' value to 'Metric' if it contains 'MEASURED IN'
    Item = ifelse(grepl("MEASURED IN", Item), NA, Item) # Set 'Item' to NA where we moved the value
  )
#Remove 'production' to its correct way.
strawberry <- strawberry |>
  mutate(
    Item = ifelse(grepl("PRODUCTION", Metric), "PRODUCTION", Item), # Move 'PRODUCTION' to 'Item'
    Metric = ifelse(grepl("PRODUCTION", Metric), sub("PRODUCTION, ", "", Metric), Metric)  # Remove 'PRODUCTION' from 'Metric'
  )
#Remove 'utilized' from category to Item
strawberry <- strawberry |>
  mutate(
    Item = ifelse(grepl("UTILIZED", Category, ignore.case = TRUE), 
                  paste("UTILIZED", Item, sep = " "), # Combine 'Item' with 'Utilized'
                  Item),  # Keep 'Item' unchanged if 'Utilized' not found
    Category = ifelse(grepl("UTILIZED", Category, ignore.case = TRUE), NA, Category)# Set 'Category' to NA where 'Utilized' is moved
  )
#Consider a better waty to move items in one step.
movingitem<- c("ACRES BEARING", "ACRES NON-BEARING", "ACRES GROWN", "OPERATIONS WITH AREA BEARING", "YIELD", "ACRES HARVESTED", "ACRES PLANTED", "OPERATIONS WITH AREA GROWN", "OPERATIONS WITH AREA NON-BEARING","PRODUCTION")

# Move terms from 'Metric' or 'Category' to 'Item' without replacing 'Metric' data
strawberry <- strawberry |>
  mutate(Item = ifelse(grepl(paste(movingitem, collapse = "|"), Category, 
                             ignore.case = TRUE) & is.na(Item), Category, 
      ifelse(grepl(paste(movingitem, collapse = "|"), Category, ignore.case = TRUE), 
             paste(Item, Category, sep = ", "), Item)
    ),
    Category = ifelse(grepl(paste(movingitem, collapse = "|"), Category, 
                            ignore.case = TRUE), 
                      NA, Category)
  )

```

#Fixing the leading space
```{r}
strawberry$Category[1]
strawberry$Item[2]
strawberry$Metric[6]
strawberry$Domain[1]
#Trim the white space

strawberry$Category <- str_trim(strawberry$Category, side = "both")
strawberry$Item <- str_trim(strawberry$Item, side = "both")
strawberry$Metric <- str_trim(strawberry$Metric, side = "both")
```

#Split both 'Domain' and 'Domain Category' columns
I find that in domain and domain category, the information is complicated, in this step I will split them into columns just like I did in 'Data Item'.
```{r}
# Split the Domain column into multiple categories
strawberry <- strawberry |>
  separate_wider_delim(
    cols = Domain,          
    delim = " , ",          
    names = c("Area Grown", "Fertilize", "Organic", "Chemical"),
    too_many = "merge",     
    too_few = "align_start" 
  )

#Loading variables to each column
strawberry <- strawberry |>
  mutate(
    Chemical = ifelse(grepl("CHEMICAL", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
    Organic = ifelse(grepl("ORGANIC", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
    Fertilize = ifelse(grepl("FERTILIZER", `Area Grown`, ignore.case = TRUE), `Area Grown`, NA),
    `Area Grown`= ifelse(grepl("CHEMICAL|ORGANIC|FERTILIZER", `Area Grown`, ignore.case = TRUE), NA, `Area Grown`)
  )

#Dealing with 'Domain Category' column
strawberry <- strawberry |>
  mutate(
    Chemical = ifelse(grepl("CHEMICAL", `Domain Category`, ignore.case = TRUE), 
                      `Domain Category`, 
                      Chemical),

    Organic = ifelse(grepl("ORGANIC", `Domain Category`, ignore.case = TRUE), 
                     `Domain Category`, 
                     Organic),
    
    Fertilize = ifelse(grepl("FERTILIZER", `Domain Category`, ignore.case = TRUE), 
                     `Domain Category`, 
                     Fertilize),

    `Area Grown` = ifelse(grepl("AREA", `Domain Category`, ignore.case = TRUE), 
                     `Domain Category`, 
                     `Area Grown`),
    
    `Domain Category` = ifelse(grepl("CHEMICAL|ORGANIC|FERTILIZER|AREA", `Domain Category`, ignore.case = TRUE), NA, `Domain Category`)
  )

#Move 'Total' to its best place
strawberry <- strawberry |>
  mutate(Item = ifelse(grepl("Total", `Area Grown`, ignore.case = TRUE), 
                  paste("Total", Item, sep = " "), 
                  Item),

    `Area Grown` = ifelse(grepl("Total", `Area Grown`, ignore.case = TRUE), NA, `Area Grown`)
  ) 

```

#Split Chemical into three renamed columns
Now we have splited chemical information, my next step is splitting it into 3 columns.
```{r}
table(strawberry$Chemical)
strawberry <- strawberry |>
  mutate(Chemical = str_replace_all(Chemical, "[,:=()]", ","))

#Split it into three columns
strawberryc<- strawberry |>
  separate_wider_delim(
    cols = Chemical,
    delim = ",",
    names = c("Type", "Ingredient", "Code"), #Separate Chemical into type, ingredient, and code.
    too_many = "merge",
    too_few = "align_start"
  )  
#Filling in the columns
strawberryc<- strawberryc |>
  mutate(
    Type = ifelse(Type == "CHEMICAL" | is.na(Type), Ingredient, Type),
    
    Ingredient = ifelse(!is.na(Ingredient), str_extract(Code, "\\b[A-Za-z\\-\\.\\s]+\\b"), Ingredient), #"\\b[A-Za-z0\\-\\.\\s]+\\b" are regular expressions, which are used to extract specific numbers or words
    
    Code = str_replace(Code, "\\b[A-Za-z\\-\\.\\s]+\\b", "")
  )

#Clean 'Code' Column
strawberryc <- strawberryc |>
  mutate(
    Code = str_replace_all(Code, "^\\s*,+|,+\\s*$|\\s*,\\s*,+", ""),
    Code = str_trim(Code)  
  )

head(strawberryc)
```